Does It Scale? Symbolic Execution of PHP Web Applications

### 2017-01-18 paper.tex
refactored 3 mainly, integrated tiens feedback

\usepackage{tikz}
\usetikzlibrary{patterns}

@@

\usepackage[zerostyle=d]{newtxtt}
%\renewcommand*\familydefault{\ttfamily} %% Only if the base font of the document is to be typewriter style
\usepackage[T1]{fontenc}

@@

\lstset{
	language=PHP,
	extendedchars = \true,
	basicstyle=\ttfamily\footnotesize,
	basewidth={0.5em,0.4em}, % for other columns settings
	commentstyle = \color{gray},
	columns=fullflexible,
	numbers=left,
	numberstyle=\tiny,
	frame=tb,
	showstringspaces=false,
	xleftmargin=2em,
	framexleftmargin=1.5em,
	backgroundcolor=\color{lightgrey},
	inputencoding = utf8x,
    keepspaces = true,
    keywordstyle = \bfseries
}

@@

\usepackage{pgfplots, pgfplotstable}

\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}

@@

% DOI
\acmDOI{10.475/123_4}

% ISBN
\acmISBN{123-4567-24-567/08/06}

%Conference
\acmConference[ISSTA'17]{ACM Woodstock conference}{July 2017}{Santa Barbara,
California, USA}
\acmYear{1997}
\copyrightyear{2016}

\acmPrice{15.00}

\begin{document}
\title{Does It Scale? Static Output Approximation of PHP Web Applications Using
Symbolic Execution}
\titlenote{Produces the permission block, and copyright information}
%\subtitle{Extended Abstract}
%\subtitlenote{The full version of the author's guide is available as  \texttt{acmart.pdf} document}

@@

\begin{abstract}
Dynamic web applications have become widely popular and are to
a large proportion based on the scripting language PHP. Output approximation of web
applications enables a range of additional tool support as well as possibilities
for vulnerability detection. Unfortunately, recent approximation approaches have
only been evaluated for smaller systems.

This paper presents an experience report about the scalability of output
approximation using symbolic execution of state-of-the-art PHP web
applications. For a symbolic execution engine extended with support for
object-oriented programming and arrays, we identified language features and
corresponding programming patterns that impede symbolic execution and limit the
scalability of this approach. Our findings include: (1) Dynamic features such as
functions and includes are prone to fail for certain programming patterns. (2)
Expressions containing elements from I/O, databases or files can heavily impede
symbolic execution. Our findings provide useful guidelines to design new
tools and also to improve the development process of statically analyzable web
applications.
\end{abstract}

@@

\section{Introduction}
With the emerging world wide web, dynamic web applications have become
widely popular. Various different implementation techniques have emerged,
ranging from technologies for programming languages (JSP, ASP .NET) and web
application frameworks for  script languages (Ruby On Rails, Django) to
languages tailored specifically to the domain of web applications. PHP
cite{phpNET} is a programming language focused on server-side application
development. As of 2012, it was used by 78.8 percent of the ten million most popular websites (according
to Alexa popularity ranking) \cite{alexaPHP}, ranking 7th on the TIOBE
programming community index \cite{tiobePHP}, and was the ranked as the 6th
most popular language on GitHub \cite{githubPHP}.

@@

Knowledge about different output output variants is leveraged by a number of
analyses for the domain of web applications. Previous work presented tools
for detecting and locating HTML validation errors
\cite{Nguyen:2011:AFH:2190078.2190142}, computing program slices across
server-side and client-page code \cite{Nguyen:2015:CPS:2786805.2786872}, or
easing development and maintenance by extending IDE support for web applications
with code navigation
\cite{Nguyen:2015:VIS:2819009.2819140,Nguyen:2014:BCG:2635868.2635928}.

@@

To investigate the question, whether we can have a practical and scalable
symbolic executoin engine, we re-implemented the engine with the specifications
of the previous symbolic execution semantics \cite{Nguyen:2014:BCG:2635868.2635928} for PHP and additional features, such
as object-oriented programming. We evaluate our
symbolic execution engine for large-scale and modern PHP systems.

During the introduction of new semantics, we addressed two  trade-offs between
accuracy of our e xecution and performance: For method calls with ambiguous
targets as well as concrete execution of loops the number of program states can
become infeasible. This may require additional engineering adjustments to tame
the state space explosion at cost of accuracy.

Based on out observations we identify conceptual limitations of symbolic
execution for PHP.  Dynamic features such as indirect function calls and
include expression require concrete information to be evaluated properly. Since
these expressions may be assembled at runtime and can contain symbolic
information, for many cases this restricts our engine from further execution.
Although some effort to statically approximate include expression can be spent,
it is often non-trivial since expressions can contain information coming from
various inputs including databases, user inputs or configuration files. We
empirically evaluate our observations for a corpus including large PHP systems that are still maintained.

Our key contribution in this paper include (1)  a new tool infrastructure to
statically approximate the client-page output for PHP web applications using
symbolic execution and (2) a report of out observations for state-of-the-art
web applications as well as an empirical evaluation of conceptual limitations
of symbolic executions for PHP web applications.

\section{State Of the Art}\label{sec:stateoftheart}
Symbolic execution is an analysis techniques used by the tools
whose practicality we aim to evaluate. This section recaps the idea of symbolic
execution, the tools which make use of it, and both conceptual and
language-related limitations.

\subsection{Symbolic Execution}
Symbolic execution is a static analysis technique that was proposed by James C.
King in 1976 \cite{King1976}. Symbolic execution allows to explore feasible
paths in a program. Based on normal program execution, the execution semantics
is extended to handle symbolic values. An execution becomes symbolic by
assuming symbolic values as program inputs rather than obtaining concrete
values \cite{King1976,Darringer1978}. These symbolic values are then propagated
and used throughout the execution.

@@

\begin{figure}[t]
	\centering
	\begin{subfigure}[center]{0.48\textwidth}%%[multicols=1]
\begin{lstlisting}
<form>
  <input type="text" name="name">
  <input type="submit" value="Submit">
</form>

@@

$greeting = "Hello " . $_POST['name'] .  "!";
if (!isset($_POST['name'])) {
  echo "No name entered!";
} else {
  echo $greeting;
}
?>
\end{lstlisting}
		\caption{Snippet of HTML and PHP server-side web application code.
		\label{fig:web_application}}
    \end{subfigure}

@@

    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[center]{0.48\textwidth}
\begin{lstlisting}[mathescape]
<form>
  <input type="text" name="name">
  <input type="submit" value="Submit">
</form>

@@

\end{lstlisting}
		\caption{
			Approximation client paege output of the server-side web application code in
			Figure \ref{fig:web_application}. Variability as well as repetition annotations
			are represented by preprocessor directives.
			\label{fig:output_example}
		}
    \end{subfigure}
    \vspace{2mm}
    \caption{
    	Illustration of static ouptut approximation using symbolic
    	execution
    	\label{fig:code_example}
    }
\end{figure}

@@

The following symbolic execution semantics for PHP features based on
previous work \cite{Nguyen:2014:BCG:2635868.2635928} illustrate the necessary
modifications and extensions made to handle the level of ambiguity that comes with handling symbolic values. In addition to normal output, the
output approximation contains preprocessor directives to express
repeatable parts as well as output variants. All approximated client page
output corresponding to Figure \ref{fig:web_application} consisting of two different variants
is illustrated in Figure \ref{fig:output_example}.
First, for loops in the previous execution semantics
\cite{Nguyen:2014:BCG:2635868.2635928} only one iteration is executed instead of
an arbitrary number of iterations. Since a loop condition can contain symbolic
values, it might not be feasible to determine whether, and if so, when a loop
execution terminates. Hence, any output constructed during the single loop
iteration is highlighted (\texttt{repeat \ldots endrepeat}) to possibly be repeated an arbitrary number of times. The output in line 7 is marked to be repeated an arbitrary number of times, as the number of loop iterations cannot be determined statically. Second, a similar procedure
applies to recursive functions. Once called, a recursive function is  also only
executed to recursion depth one per call, and any subsequent invocation of that
function returns a symbolic value. This, again, is due to the difficulty to
determine the exact recursion depth similar to the number of loop iterations.
Third, any form of user interaction providing input data as well as any form of input referred
to from the deployment context of the web application, i.e. configuration files
or databases, is assumed to be symbolic. The context variable
\texttt{\$\_POST} represents input submitted to the server and, hence, is
symbolic as every program input is per definition substituted by symbolic values. Finally, the
output in lines 11 and 13 respectively are marked as alternative variants
(\texttt{if \ldots else \ldots endif}) as symbolic execution of lines 18 – 22
in Figure \ref{fig:web_application} will explore both different branches.

\subsection{Existing Tool Support for PHP}\label{sec:existing_tools}
All tools proposed so far leverage a representation of all possible HTML client
page output. Since any single client page output variant can be analyzed this
enables tool support addressing the web application as a whole in spite of its
staged nature, where parts of the web application are
generated dynamically.
Based on the approximated output representation, subsequent analyses can be
conducted:
\begin{itemize}
	\item \emph{PhpSync}: Using all variants of  HTML client page output, every
	single one can be statically checked for markup validity, i.e., if the web page
	conforms to syntactical specifications for HTML and other client-side
	languages. After the tool detects validation errors, auto-fixes
	can be provided. Otherwise, it traces validation errors back to
	source code responsible for the defect, which then can be refactored manually
	\cite{Nguyen:2011:AFH:2190078.2190142}.

@@

	\item \emph{Varis}: Editor services such as “jump to declaration” are
	nontrivial for dynamic web applications due to their staged nature. Varis
	provides editor services across stages for client-side code: Starting with an
	ouput approximation a callgraph with conditional edges is constructed, which
	allows navigation in client-site code (HTML, JS and CSS) although it is
	embedded in server-side code
	\cite{Nguyen:2015:VIS:2819009.2819140,Nguyen:2014:BCG:2635868.2635928}.

\end{itemize}

@@

\section{Scalable Output Approximation} \label{sec:3}
The main objective of this experience report is to investigate whether we can
have a scalable symbolic execution engine as it is the underlying technique for
static output approximation in previous work
\cite{Nguyen:2011:AFH:2190078.2190142,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140}.
Our vision is a practical symbolic execution, i.e., it computes an accurate approximation for large and modern PHP systems with reasonable time and space consumption.

Also, we are interested in the limitations of scalable symbolic execution for
PHP web applications. Limitations may either be conceptional, where symbolic
execution of PHP code, in contrast to concrete execution, is inaccurate: For
example an include expression evaluated to a symbolic value simply provides to
little information to include a script. Or, in spite of infeasible effort
required, limitations could allow an accurate symbolic execution. For example,
exhaustively exploring all feasible paths in an application may result in an
infeasible number of program states to store; nevertheless, this so called
state explosion problem may be addressed with additional engineering.

\subsection{Towards Scalable Output Approximation?}
As our goal is to have a scalable symbolic execution engine, we approach this
question by re-implementing the symbolic execution semantics described in
previous work \cite{Nguyen:2014:BCG:2635868.2635928} and extending it with
support for additional language features. We aim to support symbolic execution
for a real-world example system of reasonable size that incorporates the
missing language features described in Section \ref{sec:limitations} and is
still developed and maintained. \textsf{WordPress} is a popular open-source
Content Management System (CMS) with around \textcolor{blue}{300.000} lines of
code that provides a vivid plug-in environment and is maintained by a large
community.
For our implementation of a symbolic execution engine for PHP,
\textsf{Oak}\footnote{The implementation can be found
at \url{github.com/smba/oak}.}, we have chosen a test-driven approach with
continuous integration using regression tests based on \textsf{WordPress} and \textsf{SchoolMate} (see Section \ref{sec:experiment_setup}). Oak incorporates the existing execution semantics
of \cite{Nguyen:2014:BCG:2635868.2635928} and extends it with support for
object-oriented orientation (class definition, class instantiation and method
invocation) as well as most PHP array operations. A list of the functions
provided by the PHP standard library can be found at \cite{PHP}.
Throughout the implementation we did not achieve high code coverage for
\sf{WordPress}. In the following we present our observations we made thorughout
the implementation regarding issues impeding symbolic execution.

\subsection{Experience Report}\label{sec:experience_report}
We identified trade-offs between accuracy and performance that may
cause the approximation to be impractical, yet possible. Additionally, we
detected limitations impeding symbolic execution conceptually: For certain
language features, concrete information rather than symbolic values may be required to
actually be executed properly as this defective execution can impact subsequent
parts of the program which may become inaccessible and may be missed.

\subsubsection{Symbolic vs Concrete Execution}
Throughout the implementation, we basically followed the symbolic execution
semantics described in previous work \cite{Nguyen:2014:BCG:2635868.2635928}.
Nevertheless, for additional language features we experienced it to be beneficial to the
approximation to execute them as precisely as possible.
For instance, \sf{WordPress} uses a text filter mechanism, where different
filter functions are applied sequentially to a string literal. Filters can
be linked to a hook using  associative arrays so that when the hook is
triggered, all filters that are linked to it are applied. As described in
Section \ref{} we execute loops only once due to an possibly
indeterminable number of iterations. PHP yet offers more loop constructs: the
do-while loop and the foreach loop. While for the do-while loop as well as for
the while loop the number of iterations is possibly indeterminable, the foreach
loop iterates over all items of a given array whose number of items is known at
runtime. For \sf{WordPress}, we are able to achieve a more accurate
approximation and higher code coverage results by concretely executing foreach
loops as more server-side code (filter functions in our case) are accessed and
executed. In turn, exhaustively executing foreach loops and filter functions
concretely is a trade-off between accuracy and performance, as a single hook
call site can be called from various contexts since we are exploring multiple
execution paths.

An additional aspect on the trade-off between concrete and symbolic execution,
or effort and accuracy respectively, is the extent to which program inputs are
assumed to be symbolic. As shown in Section \ref{} introducing symbolic values
can be problematic if they are propagated to language features requiring
concrete information, such as include expressions or indirect function calls.
We manually scaled the number of sources of symbolic inputs by providing static
dummy inputs such as a concrete form input, file name or configuration entry.
For specific cases, where concrete information is inevitably required, we can
achieve higher code coverage and a more accurate approximation, as more code
becomes accessible. Nevertheless, while detecting dynamic features can be
automated, it requires manual inspection and understanding of the web
application to provide appropriate dummy input data. Again, this is a trade-off
between the effort to provide input data and accuracy of the approximation and
code coverage respectively.

\subsubsection{Multi-Target Method Calls}
A non-conceptual problem we encountere is related to object-oriented programming. While exploring
different execution paths when the control flow is split, a variable can have
different values depending on the path condition under which a value has been
assigned to it (see the code example in Figure \ref{fig:output_example}).
Similarly, a variable can point to different objects depending on the path
condition as the snippet in Figure \ref{fig:multi_target} illustrates. The variable \tt{\$person} can either
point to object \tt{Alice} or \tt{Bob}. In addition, for any method invocated on
that variable the actual target is ambiguous as there are two, depending on the
path condition. In order to have a sound symbolic execution, both method calls
need to be executed. Nevertheless, the number of contexts for which a method can
be called can grow rapidly, and easily become infeasible to execute
exhaustively.

We implemented several heuristic approaches to tame this state space explosion
for multi-target method calls, including executing a method only for a subset of
target objects, or just a single object. This is a trade-off between accuracy
and soundness of the symbolic execution and performance time required.

\subsubsection{Dynamic Language Features}
\label{sec:experiment_dynamicfeaturess} 
Conceptually, we identified two language features for which symbolic execution
failed frequently. As stated in Section \ref{sec:workflow}, dynamic web
applications are staged and client page output is computed at runtime. Symbolic
execution has shown to be a useful technique to approximate dynamically
generated output in theory, though it is challenged by language features that
require concrete information rather than symbolic information at runtime. In particular, this
applies to language features that provide access to reusable code of the web
application: include expressions and function calls.

First, include expressions in PHP usually require a string value representing
the path to the script file or template to include. If this string value is
statically provided by a string literal or can be constructed unambiguously, the
symbolic execution engine was able to resolve the include expression and include
the desired file. However, it is also common to assemble include expressions not
only from static string literals, but from several sources including user input
information, database query results or configuration files. Figure
\ref{fig:dynamic_includes} shows a small example of a dynamic include containing
information retrieved from a database. Since any input for a symbolically
executed application is symbolic this information can be propagated to include
expressions. Also, configuration files represent only one particular
configuration or even just default values or placeholders. Hence, symbolic
execuion is prone to fail for dynamic includes if symbolic information is contained.

@@

\subsubsection{Interpreter Customization}
We have approached issues we encountered, including
multiple-target method calls and foreach loops, in different ways as these
represent a trade-off between accuracy of execution and performance.
For method calls having ambiguous targets, we implemented different modes of
execution, ranging from a complete mode where a method is executed for all
possible targets to abstract modes with either a subset of targets or only a
single target. In the latter cases where targets are discarded, we assumed
symbolic values as method return values. Our symbolic interpreter can be
customized to work in the desired mode to scale accuracy or performance.
Nevertheless, for \sf{WordPress} these variations did not have significant
impact on the output approximation as symbolic execution was rather impeded by
dynamic features.
Also, for foreach loops we are able to toggle between a single iteration mode
and a complete iteration mode. Although we were able to increase code coverage
by accessing more code, it did not significantly increase our output
approximation.
Note that these variants have only been tested for \sf{WordPress} to better
understand possible trade-offs or simply achieve any output approximation in a
reasonable period of time. We believe based on our experience dynamic features
to be the main conceptual limitation impeding code coverage as well as output
approximation quality. In the next Section we analyze in detail the impact of
dynamic features with concrete foreach loop execution and complete
multiple-target execution for method calls.

>>>>

\usepackage{tikz}
	\usepackage{pgf, pgfplots, pgfplotstable}
	\usetikzlibrary{arrows,automata}
	\usetikzlibrary{patterns}

@@

\usepackage[zerostyle=d]{newtxtt}
\usepackage[T1]{fontenc}

@@

\lstset{
	language=PHP,
	extendedchars = \true,
	basicstyle=\ttfamily\footnotesize,
	basewidth={0.5em,0.4em}, % for other columns settings
	commentstyle = \color{black},
	columns=fullflexible,
	numbers=left,
	numberstyle=\tiny,
	frame=tb,
	showstringspaces=false,
	xleftmargin=2em,
	framexleftmargin=1.5em,
	backgroundcolor=\color{lightgrey},
	inputencoding = utf8x,
    keepspaces = true,
    keywordstyle = \bfseries
}

@@

@@

% DOI
%\acmDOI{10.475/123_4}

% ISBN
%\acmISBN{123-4567-24-567/08/06}

%Conference
\acmConference[ISSTA'17]{ACM Woodstock conference}{July 2017}{Santa Barbara,
 California, USA}
%\acmYear{1997}
%\copyrightyear{2016}

%\acmPrice{15.00}

\begin{document}
\title{Does It Scale? Symbolic Execution of PHP Web Applications}
%\titlenote{Produces the permission block, and copyright information}
%\subtitle{Extended Abstract}
%\subtitlenote{The full version of the author's guide is available as  \texttt{acmart.pdf} document}

@@

\begin{abstract}
\textcolor{gray}{
Dynamic web applications have become widely popular and are to
a large proportion based on the scripting language PHP. Output approximation of web
applications enables a range of additional tool support as well as possibilities
for vulnerability detection. Unfortunately, recent approximation approaches have
only been evaluated for smaller systems.}
\textcolor{gray}{This paper presents an experience report about the scalability of output
approximation using symbolic execution of state-of-the-art PHP web
applications. For a symbolic execution engine extended with support for
object-oriented programming and arrays, we identified language features and
corresponding programming patterns that impede symbolic execution and limit the
scalability of this approach. Our findings include: (1) Dynamic features such as
functions and includes are prone to fail for certain programming patterns. (2)
Expressions containing elements from I/O, databases or files can heavily impede
symbolic execution. Our findings provide useful guidelines to design new
tools and also to improve the development process of statically analyzable web
applications.}
\end{abstract}

@@

\section{Introduction}
With the emerging world wide web, dynamic web applications have become
widely popular. Various different implementation techniques have emerged,
ranging from technologies for programming languages (JSP, ASP .NET) and web
application frameworks for  script languages (Ruby On Rails, Django) to
languages tailored specifically to the domain of web applications. PHP
\cite{phpNET} is a programming language focused on server-side application
development. As of 2012, it was used by 78.8 percent of the ten million most popular websites (according
to Alexa popularity ranking) \cite{alexaPHP}, ranking 7th on the TIOBE
programming community index \cite{tiobePHP}, and was the ranked as the 6th
most popular language on GitHub \cite{githubPHP}.

@@

Knowledge about different output variants is leveraged by a number of
analyses for the domain of web applications. Previous work presented tools
for detecting and locating HTML validation errors
\cite{Nguyen:2011:AFH:2190078.2190142}, computing program slices across
server-side and client-page code \cite{Nguyen:2015:CPS:2786805.2786872}, or
easing development and maintenance by extending IDE support for web applications
with code navigation
\cite{Nguyen:2015:VIS:2819009.2819140,Nguyen:2014:BCG:2635868.2635928}.

@@

To investigate the question, whether we can have a practical and scalable
symbolic executoin engine, we re-implemented the engine with the specifications
of the previous symbolic execution semantics \cite{Nguyen:2014:BCG:2635868.2635928} for PHP and additional features, such
as object-oriented programming. We evaluate our
symbolic execution engine for several large-scale and modern PHP systems.

During the introduction of new semantics, we addressed two  trade-offs between
accuracy of our execution and performance: For method calls with ambiguous
targets as well as concrete execution of loops the number of program states can
become infeasible. This may require additional engineering adjustments to tame
the state space explosion at cost of accuracy.

Based on observations we identify conceptual limitations of symbolic
execution for PHP.  Dynamic features such as indirect function calls and
include expression require concrete information to be evaluated properly. Since
these expressions may be assembled at runtime and can contain symbolic
information, for many cases this restricts a symbolic execution engine from
further execution.
Although some effort to statically approximate include expression can be spent,
it is often non-trivial since expressions can contain information coming from
various inputs including databases, user inputs or configuration files. We
empirically evaluate our observations for a corpus including large, actively maintained PHP systems.

Our key contribution in this paper include (1)  a new tool infrastructure to
statically approximate the client-page output for PHP web applications using
symbolic execution and (2) a report of observations for state-of-the-art
web applications as well as an empirical evaluation of conceptual limitations
of symbolic executions for PHP web applications.

\section{State Of the Art}\label{sec:stateoftheart}
This section recaps the idea of symbolic
execution, the tools which make use of it, and both conceptual and
language-related limitations.

\subsection{Symbolic Execution}
Symbolic execution is a static analysis technique that was proposed by King
\cite{King1976} in 1976. Symbolic execution allows to explore feasible paths in a program. Based on normal program execution, the execution semantics
is extended to handle symbolic values. An execution becomes symbolic by
assuming symbolic values as program inputs rather than obtaining concrete
values \cite{King1976,Darringer1978}. These symbolic values are then propagated
and used throughout the execution.

@@

\begin{figure}[t]
	\centering
	\begin{subfigure}[center]{0.48\textwidth}%%[multicols=1]
\begin{lstlisting}
<form method="POST">
  <input type="text" name="name">
  <input type="submit" value="Submit">
</form>

@@

$greeting = "Hello " . $_POST['name'] .  "!";
if (!isset($_POST['name'])) {
  echo "No name entered!";
} else {
  echo $greeting;
}
?>
\end{lstlisting}
		\caption{Snippet of server-side HTML and PHP web application code.
		\label{fig:web_application}}
    \end{subfigure}

@@

    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[center]{0.48\textwidth}
\begin{lstlisting}[mathescape]
<form method="POST">
  <input type="text" name="name">
  <input type="submit" value="Submit">
</form>

@@

\end{lstlisting}
		\caption{
			Approximated client page output of the server-side code in Figure
			\ref{fig:web_application}. The directives in lines 6 and 8 highlight output
			that can be repeated an arbitrary number of times, the directives in line 10
			and 14 mark two alternative output parts.
			\label{fig:output_example}
		}
    \end{subfigure}
    \vspace{2mm}
    \caption{
    	Illustration of static ouptut approximation using symbolic
    	execution
    	\label{fig:code_example}
    }
\end{figure}

@@

The following symbolic execution semantics for PHP features based on
previous work \cite{Nguyen:2014:BCG:2635868.2635928} illustrate the necessary
modifications and extensions made to handle the level of ambiguity that comes with handling symbolic values. In addition to normal output, the
output approximation contains preprocessor directives to express
repeatable parts as well as output variants. All approximated client page
output corresponding to Figure \ref{fig:web_application} consisting of two different variants
is illustrated in Figure \ref{fig:output_example}.

First, for loops in the previous execution semantics
\cite{Nguyen:2014:BCG:2635868.2635928} only one iteration is executed instead of
an arbitrary number of iterations. Since a loop condition can contain symbolic
values, it might not be feasible to determine whether, and if so, when a loop
execution terminates. Hence, any output constructed during the single loop
iteration is highlighted (\texttt{repeat \ldots endrepeat}) to possibly be
repeated an arbitrary number of times. The output in line 7 is marked to be
repeated an arbitrary number of times, as the number of loop iterations cannot
be determined statically.

Second, a similar procedure applies to recursive
functions. Once called, a recursive function is  also only executed to recursion
depth one per call, and any subsequent invocation of that function returns a
symbolic value. This, again, is due to the difficulty to determine the exact
recursion depth similar to the number of loop iterations.

Third, any form of user interaction providing input data as well as any form of input referred
to from the deployment context of the web application, i.e. configuration files
or databases, is assumed to be symbolic. The context variable
\texttt{\$\_POST} represents input submitted to the server and, hence, is
symbolic as every program input is per definition substituted by symbolic values. Finally, the
output in lines 11 and 13 respectively are marked as alternative variants
(\texttt{if \ldots else \ldots endif}) as symbolic execution of lines 18 – 22
in Figure \ref{fig:web_application} will explore both different branches.

\subsection{Symbolic-Execution-based Tools For PHP}\label{sec:existing_tools}
All tools proposed so far leverage a representation of all possible HTML client
page output. Since any single client page output variant can be analyzed this
enables tool support addressing the web application as a whole in spite of its
staged nature, where parts of the web application are
generated dynamically.
Based on the approximated output representation, subsequent analyses can be
conducted:
\begin{itemize}
	\item \emph{PhpSync}: Using all variants of  HTML client page output, every
	single one can be statically checked for markup validity, i.e., if the web page
	conforms to syntactical specifications for HTML and other client-side
	languages. After the tool detects validation errors, auto-fixes
	can be provided. Otherwise, it traces validation errors back to
	source code responsible for the defect, which then can be refactored manually
	\cite{Nguyen:2011:AFH:2190078.2190142}.

@@

	\item \emph{Varis}: Editor services such as “jump to declaration” are
	nontrivial for dynamic web applications due to their staged nature. Varis
	provides editor services across stages for client-side code: Starting with an
	ouput approximation a callgraph with conditional edges is constructed, which
	allows navigation in client-site code (HTML, JS and CSS) although it is
	embedded in server-side code
	\cite{Nguyen:2015:VIS:2819009.2819140,Nguyen:2014:BCG:2635868.2635928}.
	\item \textcolor{gray}{Automated Web Application Testing Using Seacrch
	Software Engineering} \cite{Alshahwan2011}
\end{itemize}

@@

\section{Scalable Output Approximation?} \label{sec:3}
The main objective of this experience report is to investigate whether we can
build a scalable symbolic execution engine as it is the underlying technique
for static output approximation in previous work
\cite{Nguyen:2011:AFH:2190078.2190142,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140}.
Our vision is a practical symbolic execution, i.e., it computes an accurate
approximation for large and modern PHP systems with reasonable time and space consumption.
We approach this goal by re-implementing the symbolic execution semantics  of
\sf{PhpSync} \cite{Nguyen:2014:BCG:2635868.2635928} and extending it with support for additional language
including object-oriented programming and array operations. We aim to support symbolic
execution for a real-world example system of reasonable size that incorporates
the missing language features described in Section \ref{sec:limitations}.
\textsf{WordPress} is a popular open-source Content Management System (CMS)
with around 300.000 lines of code that provides a vivid
plug-in environment and is used.  The full implementation of \sf{Oak}, the test
suite as well as or test corpus (see Section \ref{sec:experiment_setup}) can be found at
\url{http://www.github.com/smba/oak}.

For our implementation of a symbolic execution engine for PHP, Oak, we have
chosen a test-driven approach with continuous integration using regression
tests based on \sf{WordPress} and \sf{SchoolMate} (see Section
\ref{sec:experiment_setup}). For \sf{SchoolMate}, a rather straightforward and
simple system, which we chose to compare our approximation results to those of
\sf{PhpSync}, we could achieve similar output approximation results. For
\sf{WordPress} though, we did not achieve high code coverage or an accurate
approximation for \sf{WordPress} as we faced a number of obstacles hindering
symbolic execution in different ways: The following Subsection
\ref{sec:experience_report} documents our design choices we made, apparent trade-offs we recognized, and limitations
we faced on our way towards a scalable symbolic execution engine; and what we
have learned from bypassing limitations faced in spite of manual effort for
future work. Subsection \ref{sec:experience_results} compares our observations
and results to \emph{PhpSync} and presents the motivation for our empirical evaluation of our
tool symbolic execution engine in Section \ref{sec:evaluation}.

\subsection{Experience Report}\label{sec:experience_report}

\subsubsection{Symbolic/Concrete Loop Execution}
Throughout the implementation we basically followed the symbolic execution
semantics of \sf{PhpSync} \cite{Nguyen:2014:BCG:2635868.2635928}. As described
in Section \ref{sec:workflow} for loops, only one iteration is executed since
the number of iterations is indeterminable for most cases. While this is the
case for while and do-while loops, the number of loop iterations for for or
foreach loops is generally determinable and known before entering the loop
body. For loops use a loop counter that usually is incremented or decremented
after each loop iteration until a loop condition is not satisfied anymore;
foreach loops simply iterate over a given array value.
Our execution semantics follows this scheme and simulates concrete execution
for those types of loops if possible. Otherwise, for instance, if a loop
counter, loop counter limit for for loops, or variable for foreach loops is
symbolic, we resort to executing only a single iteration. This design choice
was made since it is more precise and, in contrast to only considering one
iteration in all cases, slightly increased code coverage for \sf{WordPress}.

\subsubsection{Method Target Space Explosion}
For method invocations that for concrete execution have only one target object,
exploring multiple execution paths can cause the space of target objects to
grow rapidly. A minimal example is shown in the listing in Figure
\ref{fig:multi_target} where two different objects can be assigned to variable \tt{\$person} in line 12 and
14. For the method invocation in line 16 hence, the method is executed twice in
total since we have two alternative target objects. To address this problem, we
implement program state merging for symbolic execution. For the example in
Figure \ref{fig:multi_target}, after the control flow is split in line 11 into
two branches, we merge both branches together subsequently. For both branches two different
values have been assigned to variable \tt{\$person} under different path
conditions respectively. We implement state merging inspired by work of Sen et al. 
\cite{Sen:2015:MMS:2786805.2786830} using value summaries. A value summary can be conceived as a value
type. For state merging in our example, in the value summary, each value (or
each object in our case) is associated to the path condition under that it has
been assigned to a variable. This would allow to only pass the value summary to
the method and only invocate it once, but for our execution to be sound, all
objects would need to be of the same class type. If for instance, in our
example there would be another class definition with a method named
\tt{greet} and an instance of that class is assigned to variable \tt{\$person}
in one of the two branches, the method invocation in line 16 actually needs to be
executed for two different target objects of a different type.

Therefore, for our execution engine to be sound, we were bound to execute
invocated methods for multiple targets since we cannot assert that all values
contained in summary are of the same type. Nevertheless, to scale the target
object space, in addition, we implemented execution modes only allowing
execution of a method for a subset of possible targets and assuming symbolic
return values for the remaining targets. This execution mode was necessary for
WordPress for several, but not all entry points that we studied, in order to
terminate. All subsequent analyses in this paper do not use this execution
mode, yet it is provided in out implementation.

\subsubsection{Dynamic Language Features}
\label{sec:experiment_dynamicfeatures} 
Aside from the previous issues, we identified two dynamic language features
that we were not able to execute accurately. Dynamic features in PHP are
features that take as input rather a expression that is resolved at runtime
than a static constant value. For \sf{WordPress}, we faced missing files and
functions that are included using dynamic includes and dynamic function
invocation respectively.

First, file inclusion in PHP usually is managed by statements like \tt{include},
\tt{include\_once} or \tt{require} and \tt{require\_once}. For transitive
includes, statements ending with \tt{\_once} will not include the same file
twice. The include statements make an attempt to include a file, but do not
return a warning in contrast to the require statements. Each of these
statements takes as input either a string constant (static include) or an
expression that is evaluated at runtime (dynamic include). The string that the
expression evaluates to represents an absolute or relative paths to the target
file that the interpreter finally attempts to include.
Figure \ref{fig:dynamic_includes} shows a small example of a dynamic include containing
information retrieved from a database. Since any input for a symbolically
executed application is symbolic this information can be propagated to include
expressions. Also, configuration files represent only one particular
configuration or even just default values or placeholders. Hence, symbolic
execuion is prone to fail for dynamic includes if symbolic information is contained.

@@

\subsubsection{Symbolic/Concrete Input}
In the previous subsection, we argued that dynamic includes and dynamic
function invocations hinder symbolic execution  due to imprecise resolution of
expression if symbolic information is contained. For callable code parts as
included scripts or invocated functions, we attempted to bypass this limitation
by manually providing proper concrete input.

Consider the code example in Figure \ref{fig:indirect_calls} where a function
name could not be resolved precisely due to a missing string literal that is contained from a
database query result. By manually providing either option \tt{serif} or
\tt{sserif}, or both as alternatives, we are able to increase code coverage and
the accuracy of our approximation with manual effort by making dynamic
statements static. Of course we did not provide proper input exhaustively for
every occurrence of dynamic features, yet this illustrates a basis for possible
future work towards scalable symbolic execution. This manual approach addresses
the same problem as dynamic test input generation for web applications, where
proper concrete input is assumed rather than symbolic values to increase code
coverage. We present work on this problem in Section \ref{sec:related_work}.

\subsection{Experience Results} \label{sec:experience_results}
The symbolic execution engine \sf{PhpSync} \cite{Nguyen:2014:BCG:2635868.2635928} so far does not face the
problems we documented in Section \ref{sec:experience_report}. The evaluation of
PhpSync, i.e., the PHP systems that the tool was applied to, did neither contain object-oriented
features, nor dynamic features like dynamic includes or dynamic function
invocation with symbolic values.

As the observations so far did only refer to \sf{WordPress} as a large-scale
system, we wanted to study the impeding impact of dynamic features to symbolic
execution for a wider selection of PHP systems systematically apart from the
systems that \sf{PhpSync} was applied to for two reasons. First, the usage of
dynamic features, not only of includes and function invocations, varies from
system to system \cite{hills2015variable}. Second, we were not able to obtain results for
all entry points possible for \sf{WordPress} and hence could not compare our
approximatio results to those of \sf{PhpSync} where client page output
approximation is evaluated for all entry points. Therefore, we decided to
systematically evaluate our symbolic execution engine for systems apart from
\sf{WordPress}, which is documented in the following Section
\ref{sec:evaluation}.
